The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-09-05 manuscript.Rmd
include mark's comments

Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize. The digitization of natural history collections and other biodiversity information systems can now be used to enable synthesis and highlight knowledge gaps. Here we use these data to synthesize the state of our knowledge for marine invertebrates in the continental United States. We show that current knowledge and sampling efforts are inadequate given the diversity of the fauna and the geographic extent of US. While recent high-throughput sequencing technologies promise to greatly speed up the documentation of marine species and their ranges, its value will be greatly improved by anchoring the results into existing DNA sequence data and associated vouchers. However, the best documented phyla have at most 25% of their species sequenced for the most widely used DNA barcoding gene, cytochrome oxidase I (COI). More sampling, coordinated at the scale of the US, is needed to improve our knowledge of species and their geographic ranges. This data is needed to detect the impact of climate change and other anthropogenic strains to the environment. Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.

@@

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010]. For marine organisms, because of the diversity of their complex life history strategies, their size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011]. Additionally, coordinated marine ecosystem monitoring initiatives have been lacking [@Duffy2013,@Borja2016] (but are now getting organized, see MarineGEO [@Duffy2016marinegeo], MBON, and PISCO), making it difficult to monitor changes in species distributions and abundances. While most documented global extinctions of marine species are vertebrates, many others species may have been lost unnoticed because they were not known [@Sala2006]. There is therefore a pressing need to improve our knowledge of marine systems before their functioning are irreversibly altered, and species go extinct.

@@

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize. The development of public databases for specimen records (e.g. iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know. Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico (GOM) and the Pacific Northwest (PNW), to highlight gaps in our knowledge that provide a road map for important new directions in global research. We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on biological communities require not only to get a more accurate list of the species that inhabits the oceans, but also more comprehensive data on species distributions and abundances. These goals can only be achieved with additional cost-effective sampling, and a commitment to training more taxonomists and experts that can identify and interpret the material collected.

@@

A higher proportion of the species included in the regional list for the GOM are captured in biodiversity databases than for the PNW. While `r format_output((3890 + 2663 + 1025 + 316 + 102 + 49 )/n_distinct(total_gom_records$worms_valid_name) * 100)`%  of the species included in the checklist for the GOM have records in at least one of the biodiversity database (Fig. \@ref(fig:database-coverage)A), `r format_output((352 + 292 + 281 + 235 + 76 + 38)/n_distinct(total_pnw_records$worms_valid_name)*100)`% do for the PNW  (Fig. \@ref(fig:database-coverage)B). Additionally, a higher proportion of species occur in the biodiversity databases but are not included in the regional checklist for the GOM compared with the PNW (GOM: `r format_output((1253 + 848 + 556 + 216 + 190 + 124) / n_distinct(total_gom_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")`; PNW:
`r format_output( (281 + 151 + 145 + 80 + 78 + 45) / n_distinct(total_pnw_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")`).

@@

Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases. Using iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. Additional digitization of collections, including the capture of their geographic coordinates information, are needed to improve biodiversity knowledge at regional scales. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.

Two factors are most likely to contribute to explain why some species are in databases but are not included in the regional the checklists: (1) the species have been documented for the first time since the publication of the checklist, (2) the species records were not accessible or known to the authors.

@@

The lack of tools for the authors of these checklist to have access to museum records across institutions contributes most to explain why some species were not included in the checklists. Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists. We estimate that using biodiversity databases increases the number of species included in the checklist by `r format_output(spp_not_in_lists$gom_not_in_list/n_distinct(total_gom_species$worms_valid_name)*100)`% for the Gulf of Mexico, and `r format_output(spp_not_in_lists$pnw_not_in_list/n_distinct(total_pnw_species$worms_valid_name)*100)`% for the Pacific Northwest.

@@

The marine invertebrates of the US have generally been characterized genetically at a higher level than other parts of the world: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(combined_species), big.mark=",")`) of the fauna has DNA barcodes available.  In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g., 2.6% for the Antarctic [@Grant2009]).  This proportion varies by phylum from `r format_output(stat_barcoding$min_eez_value * 100)`% for `r stat_barcoding$min_eez_phylum` to `r format_output(stat_barcoding$max_eez_value * 100)`% for `r stat_barcoding$max_eez_phylum` (Fig. \@ref(fig:proportion-barcoded)). The GOM has comparable coverage than for the EEZ, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla. This is driven in part by the active Canadian barcoding effort in British Columbia.

@@

Increasing the computing literacy and teaching data wrangling skills to staff of museums in order for them to manipulate and use the data from their collection database is equally critical. As data across institutions get incorporated and pooled into large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.

@@

This study highlights that we have too little data about too few species of marine invertebrates. We recommend that additional coordinated field expeditions are organized to sample more comprehensively the marine environment. The relation between sampling effort and diversity from museum collections makes clear that any additional samples will improve greatly our knowledge of marine biodiversity. Concerted and collaborative efforts are needed to share expertise and limited resources.

>>>>

Museum collections have been recording our knowledge about species and their geographic ranges for about 250 years, but until recently, this information was fragmented across institutions and difficult to synthesize. The digitization of natural history collections and other biodiversity information systems can now be used to enable synthesis and highlight knowledge gaps. Here we use these data to synthesize the state of our knowledge for marine invertebrates in the continental United States. We show that current knowledge and sampling efforts are inadequate given the diversity of the fauna and the geographic extent of US. While recent high-throughput sequencing technologies promise to greatly speed up the documentation of marine species and their ranges, its value will be greatly improved by anchoring the results into existing DNA sequence data and associated vouchers. However, the best documented phyla have at most 25% of their species sequenced for even the most widely used DNA barcoding gene, cytochrome oxidase I (COI). More sampling, coordinated at the national scale, is needed to improve our knowledge of species and their geographic ranges. This data is needed to detect the impact of climate change and other anthropogenic strains to the environment. Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.

@@

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. <!-- add why? -->  While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010]. For marine organisms, because of the diversity of their complex life history strategies, their size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011]. Additionally, coordinated marine ecosystem monitoring initiatives have been lacking [@Duffy2013,@Borja2016] (but are now getting organized, see MarineGEO [@Duffy2016marinegeo], MBON, and PISCO), making it difficult to monitor changes in species distributions and abundances. While most documented global extinctions of marine species are vertebrates, many others species may have been lost unnoticed because they were not known [@Sala2006]. There is therefore a pressing need to improve our knowledge of marine systems before their functioning are irreversibly altered, and species go extinct.

@@

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize. The development of public databases for specimen records (e.g. iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know. Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico (GOM) and the Pacific Northwest (PNW), to highlight gaps in our knowledge that provide a road map for important new directions in global research. We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on biological communities require not only a more accurate list of the species that inhabits the oceans, but also more comprehensive data on species distributions and abundances. These goals can only be achieved with additional cost-effective sampling, and a commitment to training more taxonomists and experts that can identify and interpret the material collected.

@@

A higher proportion of the species included in the regional list for the GOM are captured in biodiversity databases than for the PNW. While `r format_output((3890 + 2663 + 1025 + 316 + 102 + 49 )/n_distinct(total_gom_records$worms_valid_name) * 100)`%  of the species included in the checklist for the GOM have records in at least one of the biodiversity database (Fig. \@ref(fig:database-coverage)A), only `r format_output((352 + 292 + 281 + 235 + 76 + 38)/n_distinct(total_pnw_records$worms_valid_name)*100)`% do for the PNW  (Fig. \@ref(fig:database-coverage)B). Additionally, a higher proportion of species occur in the biodiversity databases but are not included in the regional checklist for the GOM compared with the PNW (GOM: `r format_output((1253 + 848 + 556 + 216 + 190 + 124) / n_distinct(total_gom_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")`; PNW:
`r format_output( (281 + 151 + 145 + 80 + 78 + 45) / n_distinct(total_pnw_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")`).

@@

Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases. Using iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within US waters. Additional digitization of collections, including the capture of their geographic coordinates information, are needed to improve biodiversity knowledge at regional scales. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.

Two factors most likely contribute to explaining why some species are in databases but are not included in the regional the checklists: (1) the species have been documented for the first time since the publication of the checklist, (2) the species records were not accessible or known to the authors.

@@

The lack of tools for the authors of these checklists to have access to museum records across institutions explains why some species were not included in the checklists. Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists. We estimate that using biodiversity databases increases the number of species included in the checklist by `r format_output(spp_not_in_lists$gom_not_in_list/n_distinct(total_gom_species$worms_valid_name)*100)`% for the Gulf of Mexico, and `r format_output(spp_not_in_lists$pnw_not_in_list/n_distinct(total_pnw_species$worms_valid_name)*100)`% for the Pacific Northwest.

@@

The marine invertebrates of the US have generally been characterized genetically at a higher level than other parts of the world: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(combined_species), big.mark=",")`) of the fauna has DNA barcodes available.  In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g., 2.6% for the Antarctic [@Grant2009]).  This proportion varies by phylum from `r format_output(stat_barcoding$min_eez_value * 100)`% for `r stat_barcoding$min_eez_phylum` to `r format_output(stat_barcoding$max_eez_value * 100)`% for `r stat_barcoding$max_eez_phylum` (Fig. \@ref(fig:proportion-barcoded)). The GOM has comparable coverage in level of DNA barcoding of the fauna to the EEZ, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla. This is driven in part by the active Canadian barcoding effort in British Columbia.

@@

Increasing the computing literacy and teaching "data wrangling" skills to staff of museums in order for them to manipulate and use the data from their collection database is equally critical. As data across institutions get incorporated and pooled into large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.

@@

This study highlights that we have too little data about too few species of marine invertebrates. We recommend that additional coordinated field expeditions are organized to sample more comprehensively the marine environment. The relation between sampling effort and diversity from museum collections makes clear that any additional samples will greatly improve our knowledge of marine biodiversity. Concerted and collaborative efforts are needed to share expertise and limited resources.
